### 2019-07-14 report/chapter01.Rnw
introduction update

Studies get more attention and are more likely to be published, read and cited if they are implying large effects. Studies with no evidence for an effect are less likely to get published. This generates a bias called ``publication bias'', which distorts the view of the evidence for an effect. Recently, publication bias has been identified as one of the major concerns in irreproducible research \citep{Bishop.2019}. The issue has been pin-pointed in clinical science by many researchers (\citet{pb.clinicalscience1}, \citet{sterne2001}). In medical research, clinical trials study the efficacy of therapies and drugs. The gold standard in such intervention studies are randomized, (placebo) controlled, trials (RCTs). Results from such clinical studies influence the treatment of patients in daily clinical practice. The results from multiple intervention studies are summarized in meta-analyses and reflect the overall evidence for a treatment effect. Summarized treatment effects and uncertainities should be more robust and reliable. Publication bias can bias the overall effect estimates from meta-analyses and eventually lead to ineffective treatments that could lead to patient harm. \\
There is extensive literature on publication bias in meta-analyses, e.g. (\citet{pb.clinicalscience.2013}, %\citet{Hopewell02}, 
\citet{grey.literature.4}, \citet{grey.literature.3}, \citet{grey.literature.2}). 
All authors agree in that the exclusion of unpublished results in meta-analyses is responsible for overestimation of treatment effects. Although there are policies that make it mandatory to make accessible all study results \citep{fda}, it is not clear if the situation has improved yet.\\
The Cochrane Organisation has specialized on systematic reviews of healthcare interventions. Researchers that write a systematic review collect data across studies, review them and try to provide up-to-date information about specific treatment efficacies \citep{cochrane.handbook}. By extensive literature scrutinization, they try to circumvent the issue of publication bias. Earlier research however suggests that the efforts are only partially succesful, and that there still is publication bias within the reviews (\citet{Egger}, \citet{Ioannidis2007}, \citet{kicinsky}, \citet{vanAert.2019}). In this publications, the results of a part of the Cochrane Database of systematic reviews is analysed with small study effect tests or bayesian hierarchical selection models \citep{kicinsky}. They all find moderate to large evidence for publication bias in the Database. A more thorough discussion of their methods and results will follow in chapter \ref{ch:Discussion}. However, none of the research so far has estimated the amount and impact of publication bias on meta-analytical findings thoroughly and with the most suitable methods. Also, the results are often presented in the form of dichotomized hypothesis tests, a practice known to contribute to publication bias itself (cite sb).\\
The aim of this thesis is to use prevailed methods to detect publication bias, and make use of the full amount of data that the Cochrane Organisation provides. At the end, an up-to-date estimate of the prevalence of publication bias in the data shall be given, as well as its implications regarding the effect sizes of the treatments. To achieve this, methods to detect and adjust for publication bias in meta-analysis are applied on the data.

Already in 1978, scientists raised concern that effect sizes could be overestimated systematically if journals were ``not willing to drop their policy to use statistical significance as one of their criteria for publication'' \citep{dunlap1978}. Scientific findings get more attention and are more likely to be published, read and cited if they are implying large effects. The issue has been pin-pointed in clinical science by many researchers (\citet{pb.clinicalscience1}, \citet{sterne2001}). The concern is that overestimated efficacy of the treatments could lead to harm to patients. \\
A measure to counter the presence of single, overestimated study results is a meta-analysis \citep{metaanalysis}, a practice originally developped in psychology. Summarized treatment effects and uncertainities should be more robust and reliable. The pitfall is that when the entire body of evidence is again a biased sample of studies with high effects, the estimate is again overestimating the true effect. 

>>>>

Studies get more attention and are more likely to be published, read and cited if they are implying large effects. Studies with no evidence for an effect are less likely to get published. This generates a bias called ``publication bias'', which distorts the view of the evidence for an effect. Recently, publication bias has been identified as one of the major concerns in irreproducible research \citep{Bishop.2019}. The issue has been pin-pointed in clinical science by many researchers (\citet{pb.clinicalscience1}, \citet{sterne2001}). In medical research, clinical trials study the efficacy of therapies and drugs. The gold standard in such intervention studies are randomized, (placebo) controlled, trials (RCTs). Results from such clinical studies influence the treatment of patients in daily clinical practice. The results from multiple intervention studies are summarized in meta-analyses and reflect the overall evidence for a treatment effect. Summarized treatment effects and uncertainities should be more robust and reliable. Publication bias can bias the overall effect estimates from meta-analyses and eventually lead to ineffective treatments that could lead to patient harm. \\
There is extensive literature on publication bias in meta-analyses, e.g. (\citet{pb.clinicalscience.2013}, %\citet{Hopewell02}, 
\citet{grey.literature.4}, \citet{grey.literature.3}, \citet{grey.literature.2}). 
All authors agree in that the exclusion of unpublished results in meta-analyses is responsible for overestimation of treatment effects. Although there are policies that make it mandatory to make accessible all study results \citep{fda}, it is not clear if the situation has improved yet.\\
The Cochrane Organisation has specialized on systematic reviews of healthcare interventions. Researchers that write a systematic review collect data across studies, review them and try to provide up-to-date information about specific treatment efficacies \citep{cochrane.handbook}. By extensive literature scrutinization, they try to circumvent the issue of publication bias. Earlier research however suggests that the efforts are only partially succesful, and that there still is publication bias within the reviews (\citet{Egger}, \citet{Ioannidis2007}, \citet{kicinsky}, \citet{vanAert.2019}). In this publications, the results of a part of the Cochrane Database of systematic reviews is analysed with small study effect tests or bayesian hierarchical selection models \citep{kicinsky}. They all find moderate to large evidence for publication bias in the Database. A more thorough discussion of their methods and results will follow in chapter \ref{ch:Discussion}. None of the research so far has estimated the amount and impact of publication bias on meta-analytical findings thoroughly and with the most suitable methods. Also, the results are often presented in the form of dichotomized hypothesis tests, a practice known to contribute to publication bias itself (cite sb).\\
The aim of this thesis is to use prevailed methods to detect publication bias, and make use of the full amount of data that the Cochrane Organisation provides. At the end, an up-to-date estimate of the prevalence of publication bias in the data shall be given, as well as its implications regarding the effect sizes of the treatments. To achieve this, methods to detect and adjust for publication bias in meta-analysis are applied on the data.

\subsection{Analysing publication bias}
There are multiple ways to assess the amount of publication bias. For instance, one can follow studies and control if they are getting published depending on their findings \citep{publication.fate}. One often finds that positive findings (\ie large effects) are reported and published more often (see also an example in social sciences: \citet{social.sciences.publication.bias}). Another way is to compare results in study registries with results published in journals (\eg \citet{pb.clinicalscience.2013}). Again one finds systematic differences between published and unpublished results. \\
A third way is to assess the so-called small study effect in a meta-analysis, that is, a systematic difference between effects of large and small studies. Evidence for small study effects can oftentimes be interpreted as evidence for publication bias \citep{Egger}. The underlying rational is that journal editors are guided by two simple rules (\citet{excess.significance}, \citet{ioannidis.2019}): 
\begin{itemize}
\item Publish new and trend-setting findings that lead to many citations, increasing the journal impact factor (JIF).
\item Publish large trials that are likely to set the standard for a given scientific question, again to increase the JIF.
\end{itemize}
This then leads to the so-called asymmetry in the funnel plot of published studies; small studies show larger effect sizes than larger studies do. By means of simple linear regression, one can investigate how much evidence there is for this asymmetry. For illustration purposes, a meta-analysis with large funnel plot asymmetry is shown %\citep{pre.eclampsia}:

\begin{figure}
<<echo = FALSE, fig.height = 3>>=
PATH_HOME = path.expand("~") # user home
PATH = file.path(PATH_HOME, 'Data/PubBias')
PATH2 = file.path(PATH_HOME, 'PubBias')
FILE = 'cochrane_2019-07-04.csv'
PATH_DATA = file.path(PATH, 'data')
PATH_CODE = file.path(PATH2, 'code')
PATH_RESULTS = file.path(PATH2, 'results_new')
PATH_FIGURES = file.path(PATH_RESULTS, 'figures')
load(file.path(PATH_RESULTS, "data_used_for_analysis.RData"))
load(file.path(PATH_RESULTS, "meta_analyses_summary_complete.RData"))

pb.ex <- 111673
require(meta)
require(tidyverse)
par(mfrow = c(1,2))
tp <- data.ext2 %>% filter(meta.id == pb.ex)
funnel(metabin(event.e = events1, event.c = events2, n.e = total1, n.c = total2, data = tp),
       comb.fixed = F, comb.random = F)

nopb.ex <- meta.f$meta.id[32]
tp <- data.ext2 %>% filter(meta.id == nopb.ex)
funnel(metabin(event.e = events1, event.c = events2, n.e = total1, n.c = total2, data = tp),
       comb.fixed = F, comb.random = F)
@
\end{figure}

The purpose of a funnel plot is to visualise if an how the estimated treatment effects converge to a common ``real'' treatment effect value (here, the weighted mean treatment effect with weights = inverse variances). This is also shown by the triangle depicted by the dotted line (the spire is at the weighted mean). \\
We see that while the studies in the meta-analysis on the left are not filling the triangle evenly, but rather accumulate at the left side, they seem to be more evenly distributed in the left triangle. From this, we would ultimately conclude that some sort of bias or heterogeneity is distorting the estimate of the overall treatment effect.\\
Importantly, this method does not assess publication bias directly. But note that publication bias is also not assessable with the other methods if no complete study registries are available. Thus, the estimation of small study effects is oftentimes the most efficient and simple way to analyse publication bias.

% Already in 1978, scientists raised concern that effect sizes could be overestimated systematically if journals were ``not willing to drop their policy to use statistical significance as one of their criteria for publication'' \citep{dunlap1978}. Scientific findings get more attention and are more likely to be published, read and cited if they are implying large effects. The issue has been pin-pointed in clinical science by many researchers (\citet{pb.clinicalscience1}, \citet{sterne2001}). The concern is that overestimated efficacy of the treatments could lead to harm to patients. \\
% A measure to counter the presence of single, overestimated study results is a meta-analysis \citep{metaanalysis}, a practice originally developped in psychology. Summarized treatment effects and uncertainities should be more robust and reliable. The pitfall is that when the entire body of evidence is again a biased sample of studies with high effects, the estimate is again overestimating the true effect. 
% % 

### 2019-07-14 report/chapter02.Rnw
Report based on the new data
introduction update

\section{Transformation between effect sizes} \label{sec:transformation.effectsizes}
Assuming that binary outcomes result from a dichotomization of originally continuous random variables, in this case, the logistic distribution, a transformation from typical binary effect measures to Cohen's $d$ can be achieved \cite[47]{Intro.meta}. \\
Let $\theta$ be a log odds ratio and $s$ it's standard error. Cohen's $d$ and it's variance $s_d^2$ is obtained by

>>>>

\section{Transformation between Effect Measures} \label{sec:transformation.effectsizes}
Assuming that binary outcomes result from a dichotomization of originally continuous random variables, in this case, the logistic distribution, a transformation from typical binary effect measures to Cohen's $d$ can be achieved \cite[47]{Intro.meta}. \\
Let $\theta$ be a log odds ratio and $s$ it's standard error. Cohen's $d$ and it's variance $s_d^2$ is obtained by

### 2019-07-14 report/chapter03.Rnw
introduction update

#Barbiturate Examples
barbi1 <- arrange(filter(data, id == "CD000033") %>% 
          select(study.name, comparison.name, outcome.name, 
                 events1, total1, events2, total2) %>% 
            slice(c(1,3)), Study) 
barbi2 <- arrange(filter(data, id == "CD000033") %>% 
          select(Study = study.name, Comparison = comparison.name,Outcome = outcome.name), Study)
barbi2[barbi2$Study == "P\303\251rez-B\303\241rcena 2008", 1] = "Perez-Barcena 2008"
#----------------------------------------------------------------------------------------------#

@@

#----------------------------------------------------------------------------------------------#
# "IV" outcome.flag meta-analyses
dt <- meta.iv %>% group_by(outcome.measure.new) %>% count
first.iv.outcome.count <- dt$n[1]
second.iv.outcome.count <- dt$n[2]
third.iv.outcome.count <- dt$n[3]
fourth.iv.outcome.count <- dt$n[4]

@@

\subsubsection{Exclusions due to Computational Errors}
There has been cases where meta-analyses could not be analysed with Copas selection model because the likelihood function could not be optimized based on the settings. It was ommitted in this case. The analysis failed when using $z$-scores, but not standardized mean differences, and in that case, it was not ommitted.\\
The specific information about the meta-analysis is: \texttt{meta.id} = 220949, \texttt{id} = ``CD009908'', \texttt{subgroup.id} = ``CMP-001.05.02''.

\subsubsection{The Analysis Dataset}
The dataset that was ultimately tested and adjusted for publication bias comprises \Sexpr{length(meta.id.vector)} meta-analyses and \Sexpr{dim(meta.data)[1]} results. The mean number of participants in the treatment group is \Sexpr{mean(meta.data$total1, na.rm = T)} vs \Sexpr{mean(data.ext2$total1, na.rm = T)} in the unrestricted dataset and the mean total number of participants \Sexpr{mean(meta.data$total1 + meta.data$total2)} vs \Sexpr{mean(data.ext2$total1 + data.ext2$total2)}. The mean publication year is \Sexpr{mean(meta.data$study.year, na.rm = TRUE)} vs \Sexpr{mean(data.ext2$study.year, na.rm = TRUE)} in the unrestricted dataset. \\
From the meta-analyses with incomplete data (\texttt{outcome.type} == \texttt{IV}}), there are \Sexpr{first.iv.outcome.count} with \texttt{outcome.measure.new} = ..., \Sexpr{second.iv.outcome.count} ...,  \Sexpr{third.iv.outcome.count} ... and \Sexpr{fourth.iv.outcome.count} ... .

>>>>

#Barbiturate Examples
barbi1 <- arrange(filter(data, id == "CD000033") %>% 
          select(study.name, comparison.name, outcome.name, 
                 events1, total1, events2, total2) %>% 
            slice(c(1,3)), study.name) 
barbi2 <- arrange(filter(data, id == "CD000033") %>% 
          select(study.name, comparison.name, outcome.name), study.name)
barbi2[barbi2$study.name == "P\303\251rez-B\303\241rcena 2008", 1] = "Perez-Barcena 2008"
#----------------------------------------------------------------------------------------------#

@@

#----------------------------------------------------------------------------------------------#
# "IV" outcome.flag meta-analyses
dt <- meta.iv %>% group_by(outcome.measure.new) %>% count %>% arrange(desc(n))
first.iv.outcome.count <- dt$n[1]
second.iv.outcome.count <- dt$n[2]
third.iv.outcome.count <- dt$n[3]
fourth.iv.outcome.count <- dt$n[4]

@@

\subsubsection{Exclusions due to Computational Errors}
There has been cases where meta-analyses could not be analysed with Copas selection model because the likelihood function could not be optimized based on the settings. It was ommitted in this case.\\
\begin{itemize}
\item Meta-analysis with \texttt{meta.id} = 93868,  \texttt{id} = ``CD004072'', \texttt{subgroup.id} = ``CMP-006.08.01''. One small meta-analysis (\texttt{se.new} = 1.02) with a risk ratio 0.072, while the remaining risk ratios were all larger than 0.45. (and the fixed effects estimate close to 1). It failed when analysed based on log risk ratios.
\item Meta-analysis with \texttt{meta.id} = 144568,  \texttt{id} = ``CD006109'', \texttt{outcome.id} = ``CMP-001.06'' and \texttt{outcome.measure} ``MD''. $z$-score based analysis failed, although funnel plot looks more or less symmetric. Small studies with very similar \texttt{var.z} with $z$-scores 0.3 and -0.1.
\item Meta-analysis with \texttt{meta.id} = 35673,  \texttt{id} = ``CD001790'', \texttt{outcome.id} = ``CMP-001.03'' and \texttt{outcome.measure} = ``PETO.OR''. Analysis failed when using Hedges $g$ transformed effects. Large studies have $g \sim 0$ while one smaller study ($\se = 0.3$) has $g = 1$ (and another with a little larger $\se$ has $g \sim 0$).
\end{itemize}
Because the number of errors is low, the meta-analyses have only been ommitted for the specific outcome measures, but are kept in the others.

\subsubsection{The Analysis Dataset}
The dataset that was ultimately tested and adjusted for publication bias comprises \Sexpr{length(meta.id.vector)} meta-analyses and \Sexpr{dim(meta.data)[1]} results. The mean number of participants in the treatment group is \Sexpr{round(mean(meta.data$total1, na.rm = T),1)} vs \Sexpr{round(mean(data.ext2$total1, na.rm = T),1)} in the unrestricted dataset and the mean total number of participants \Sexpr{round(mean(meta.data$total1 + meta.data$total2),1)} vs \Sexpr{round(mean(data.ext2$total1 + data.ext2$total2),1)}. The mean publication year is \Sexpr{mean(meta.data$study.year, na.rm = TRUE)} vs \Sexpr{mean(data.ext2$study.year, na.rm = TRUE)} in the unrestricted dataset. \\
From the meta-analyses with incomplete data (\texttt{outcome.type} == \texttt{IV}), there are \Sexpr{first.iv.outcome.count} with \\ \texttt{outcome.measure.new} = ``Risk Ratio'', \Sexpr{second.iv.outcome.count} ``Std. Mean Difference'',  \Sexpr{third.iv.outcome.count} ``Hazard Ratio'' and \Sexpr{fourth.iv.outcome.count} ``Odds Ratio'' (and 25 additional, different outcome measures).
